Demographic Species Distribution Models

### 2019-10-18 manuscript/appendix_a.Rmd
add biotic interaction text and notation; double-subscript survival notation

The transition matrix $\mathbf{A}_i$ describes the fundamental dynamics of the population at location $i$, with cells expressing the rates of transition from each stage at the previous timestep (columns) to each stage in the present time step (rows). Thus for each adult in the previous time step $F_i \cdot S_{Ai}$ juveniles are contributed to the population in the current time step: each adult survives with probability $S_{Ai}$, and produces an average of $F_i$ offspring, with the product of these giving the expected contribution per adult. For more details of matrix population models such as this, see [@caswellMatrixPopulationModels2001].

@@

\begin{align}
\mathbf{A}_i &= \begin{bmatrix} F_i \cdot S_{Ji} & F_i \cdot S_{Ai} \\ S_{Ji} & S_{Ai}  \end{bmatrix} \\ \nonumber
ln(F_i) &= \mathbf{X}_i\beta_F \\ \nonumber
logit(S_{Ai}) &= \mathbf{X}_i\beta_S \\ \nonumber
logit(S_{Ji}) &= logit(S_{Ai}) - \gamma
\end{align}

@@

\begin{align}
\Lambda_i &=  \mathbf{n}_{i, T} \\ \nonumber
\mathbf{n}_{i,t+1} &= \mathbf{A}_{i,t+1} \mathbf{n}_{i,t} \\ \nonumber
\mathbf{A}_{i,t+1} &= \begin{bmatrix} F_{i,t} \cdot S_{Ji,t} & F_{i,t} \cdot S_{Ai,t} \\ S_{Ji,t} & S_{Ai,t}  \end{bmatrix}  \\ \nonumber
ln(F_{i,t}) &= \mathbf{X}_i \beta_F + \alpha_F \textstyle\sum_{j=1}^M{\mathbf{n}_{i,t,j}}\\ \nonumber
logit(S_{Ai,t}) &= \mathbf{X}_i \beta_S + \alpha_S \textstyle\sum_{j=1}^M{\mathbf{n}_{i,t,j}} \\ \nonumber
logit(S_{Ji,t}) &= logit(S_{Ai,t}) - \gamma
\end{align}

@@

\begin{align}
y_{i,t} &\sim Bernoulli(p_{i,t}) \\ \nonumber
p_{i,t} &= 1 - e^{-\Lambda_{i,t}} \\ \nonumber
\Lambda_{i,t} &=  \mathbf{n}_{i,t} \\ \nonumber
\mathbf{n}_{i,t+1} &= \mathbf{A}_{i,t+1} \mathbf{n}_{i,t} \\ \nonumber
\mathbf{A}_{i,t+1} &= \begin{bmatrix} F_{i,t} \cdot S_{Ji,t} & F_{i,t} \cdot S_{Ai,t} \\ S_{Ji,t} & S_{Ai,t}  \end{bmatrix}  \\ \nonumber
ln(F_{i,t}) &= \mathbf{X}_{i,t} \beta_F + \alpha_F \textstyle\sum_{j=1}^M{\mathbf{n}_{i,t,j}}\\ \nonumber
logit(S_{Ai,t}) &= \mathbf{X}_{i,t} \beta_S + \alpha_S \textstyle\sum_{j=1}^M{\mathbf{n}_{i,t,j}} \\ \nonumber
logit(S_{Ji,t}) &= logit(S_{Ai,t}) - \gamma
\end{align}

@@

@@

<to do, need to package up inteegration code into greta.dynamics?>

@@

<to do, just add multiple life stages>

Need iternatactions to be via density-dependent effects, because we can't have negative elements in the matrix?

>>>>

The transition matrix $\mathbf{A}_i$ describes the fundamental dynamics of the population at location $i$, with cells expressing the rates of transition from each stage at the previous timestep (columns) to each stage in the present time step (rows). Thus for each adult in the previous time step $F_i \cdot S_{A_i}$ juveniles are contributed to the population in the current time step: each adult survives with probability $S_{A_i}$, and produces an average of $F_i$ offspring, with the product of these giving the expected contribution per adult. For more details of matrix population models such as this, see [@caswellMatrixPopulationModels2001].

@@

\begin{align}
\mathbf{A}_i &= \begin{bmatrix} F_i \cdot S_{J_i} & F_i \cdot S_{A_i} \\ S_{J_i} & S_{A_i}  \end{bmatrix} \\ \nonumber
ln(F_i) &= \mathbf{X}_i\beta_F \\ \nonumber
logit(S_{A_i}) &= \mathbf{X}_i\beta_S \\ \nonumber
logit(S_{J_i}) &= logit(S_{A_i}) - \gamma
\end{align}

@@

\begin{align}
\Lambda_i &=  \mathbf{n}_{i, T} \\ \nonumber
\mathbf{n}_{i,t+1} &= \mathbf{A}_{i,t+1} \mathbf{n}_{i,t} \\ \nonumber
\mathbf{A}_{i,t+1} &= \begin{bmatrix} F_{i,t} \cdot S_{J_{i,t}} & F_{i,t} \cdot S_{A_{i,t}} \\ S_{J_{i,t}} & S_{A_{i,t}}  \end{bmatrix}  \\ \nonumber
ln(F_{i,t}) &= \mathbf{X}_i \beta_F + \alpha_F \textstyle\sum_{j=1}^M{\mathbf{n}_{i,j,t}}\\ \nonumber
logit(S_{A_{i,t}}) &= \mathbf{X}_i \beta_S + \alpha_S \textstyle\sum_{j=1}^M{\mathbf{n}_{i,j,t}} \\ \nonumber
logit(S_{J_{i,t}}) &= logit(S_{A_{i,t}}) - \gamma
\end{align}

@@

\begin{align}
y_{i,t} &\sim Bernoulli(p_{i,t}) \\ \nonumber
p_{i,t} &= 1 - e^{-\Lambda_{i,t}} \\ \nonumber
\Lambda_{i,t} &=  \mathbf{n}_{i,t} \\ \nonumber
\mathbf{n}_{i,t+1} &= \mathbf{A}_{i,t+1} \mathbf{n}_{i,t} \\ \nonumber
\mathbf{A}_{i,t+1} &= \begin{bmatrix} F_{i,t} \cdot S_{J_{i,t}} & F_{i,t} \cdot S_{A_{i,t}} \\ S_{J_{i,t}} & S_{A_{i,t}}  \end{bmatrix}  \\ \nonumber
ln(F_{i,t}) &= \mathbf{X}_{i,t} \beta_F + \alpha_F \textstyle\sum_{j=1}^M{\mathbf{n}_{i,j,t}}\\ \nonumber
logit(S_{A_{i,t}}) &= \mathbf{X}_{i,t} \beta_S + \alpha_S \textstyle\sum_{j=1}^M{\mathbf{n}_{i,j,t}} \\ \nonumber
logit(S_{J_{i,t}}) &= logit(S_{A_{i,t}}) - \gamma
\end{align}

@@

## Species with complex demography

<illustrate that any species for which a transition matrix can be formed can be modelled with a DSDM- at least to illustrate that multiple life stages can be considered.>

@@

<to do, need to package up integration code into greta.dynamics?>

@@

Simultaneous demographic species distribution modelling of multiple species can be achieved by replacing the population transition matrices at each site $\mathbf{A}_i$ with a block-diagonal matrix comprising the population transition matrices for all the species of interest in the diagonal blocks. For a two-species model $\mathbf{A}_i$ therefore has dimension $(M1+M2)$x$(M1+M2)$, where $M1$ and $M2$ are the number of life stages for species 1 and 2 respectively. Since an individual can neither change species or produce offspring of another species, the off-diagonal blocks are all zero. Interactions between these species can be encoded in density-dependent vital rates, with the density of both conspecific and intraspecific populations having an impact.

\begin{align}
\Lambda_{i,l} &= \mathbf{N}_{l,i,T} \\ \nonumber
\mathbf{n}_{i,t+1} &= \mathbf{A}_{i,t+1} \mathbf{n}_{i,t} \\ \nonumber
\mathbf{A}_{i,t+1} &= \begin{bmatrix} \mathbf{A}_{1,i,t+1} & & \dots & & 0 \\
& \ddots & & & \\
\vdots & & \mathbf{A}_{l,i,t+1} & & \vdots \\
& & & \ddots & \\
0 & & \dots & & \mathbf{A}_{L,i,t+1} \end{bmatrix} \\ \nonumber
\mathbf{A}_{l,i,t+1} &= \begin{bmatrix} F_{l,i,t} \cdot S_{J_{l,i,t}} & F_{l,i,t} \cdot S_{A_{l,i,t}} \\ S_{J_{l,i,t}} & S_{A_{l,i,t}}  \end{bmatrix}  \\ \nonumber
ln(F_{l,i,t}) &= \mathbf{X}_i \beta_{F_l} + \textstyle\sum_{l'=1}^{L}{\alpha_{F_{l,l'}} \mathbf{N}_{l',i,t}}\\ \nonumber
logit(S_{A_{l,i,t}}) &= \mathbf{X}_i \beta_{S_l} + \textstyle\sum_{l'=1}^{L}{\alpha_{S_{l,l'}} \mathbf{N}_{l',i,t}}\\ \nonumber
logit(S_{J_{l,i,t}}) &= logit(S_{A_{l,i,t}}) - \gamma_l \\ \nonumber
\mathbf{N}_{l,i,t} &= \textstyle\sum_{j=1+(l - 1)}^{lM}{\mathbf{n}_{i,j,t}}
\end{align}

```{r biotic_interactions}

```

## Evolutionary dynamics

As in the biotic interactions, model different genotypes with a separate set of life stages, but fecundity in off-diagonal blocks gives the probability of the offspring having a different genotype to the parent. Different genotypes can therefore have different vital rates.

How to get separate abundances? intrinsic growth rate-derived abundance, multiplied by relative abundance of each of the genotypes.

\begin{align}
a
\end{align}

```{r evolutionary_dynamics}

```

### 2019-10-18 manuscript/ms.Rmd
add biotic interaction text and notation; double-subscript survival notation

The spatially-averaged fecundity and adult survival estimates from @ryuDevelopingPopulationModels2016 were used to define priors on regression coefficients for the vital rate submodels. These estimated vital rates correspond to populations at the $J$ survey stations used in the recapture analysis (station names provided in the supplemental data to @ryuDevelopingPopulationModels2016) - locations where the species occurs. I defined a prior on $\beta_F$ and $\beta_S$ such that values were *a priori* more likely if they resulted in estimates of $F_j$ and $S_{Aj}$, for each survey station location $j$, that were more consistent with the estimates in @ryuDevelopingPopulationModels2016.
First, the means and variances of these estimates were used to calculate equivalent parameters of univariate normal distributions on the link-scale of the regression submodels, resulting in the implied priors: $ln(F_j) \sim N(-0.63, 0.18)$, $logit(S_{Aj}) \sim N(1.12, 1.28)$, and $logit(S_{Jj}) \sim N(-0.05, 0.85)$.
These normally-distributed implied priors correspond exactly to multivariate normal priors on the vectors of regression coefficients, since the matrix-multiplication that maps between regression coefficients and link-scale response is an affine transformation, under which the multivariate normal distribution is closed. The parameters of the priors over $\beta_F$ are therefore given by: $\mu_F = \mathbf{X}_*^{-1}(\text{-}0.63\, \mathbf{I})$ and $\mathbf{\Sigma} = \mathbf{X}_*^{-1} (0.18\,\mathbf{I}) \mathbf{X}_*^{-T}$, where $\mathbf{X}_*$ denotes the design matrix over the $J$ survey stations, $\mathbf{X}_*^{-1}$ its Moore-Penrose pseudoinverse, $\mathbf{X}_*^{-T}$ the transpose of that pseudoinverse, and $\mathbf{I}$ an identity matrix of dimension $J$. The prior parameters $\mu_S$ and $\mathbf{\Sigma}_S$ were defined in the same way using the implied prior parameters over $logit(S_{Aj})$: 1.12, and 1.28.
Note that these priors do not provide any information on the direction of the relationships between vital rates and environmental covariates (the intercept elements of $\mu_F$ and $\mu_S$ are informed, but the prior mean for the other terms is $\approx 0$), but by providing information on plausible magnitudes of, and correlations between these parameters the prior favours regression equations that pass close to the vital rate estimates at $F_j$ and $S_j$.
The prior for $\gamma$, the logit-difference between adult and juvenile survival, is given by the difference between the implied normal priors on logit-survival of adults and juveniles, which is also normal.
In the absence of any meaningful prior information for the competition effects or detection rates, I assigned $\alpha$ an improper uniform prior on all positive values.

@@

For example, theese linear models could be applied to very large numbers of covariates and transformations thereof in combination with regularising priors to perform automatic variable selection, similarly to the popular Maxent SDM software[@phillipsModelingSpeciesDistributions2008]. Or to specify mixed-effects models to account for clustered sampling designs, or use random slopes to share environmental reponse information between species in a biotic interaction DSDM. Instead of linear models, non-linear functions could be introduced with generalisd additive models or Gaussian process models, for which there are existing greta extension packages: greta.gam and greta.gp [@cGretaGamGeneralised2019; @goldingGretaGpGaussian2019]. These can also be used to model spatial and teemporal random effects and account for drivers of species vital rates that are not captured in the measured environmental variables. Prespecified mathematical relationships or covariate values can also be included, enabling incorporation of  spatial and temporal predictions from physiological models.

@@

Bayesian inference on the DSDM for the Carolina Chickadee by MCMC took `r round(bbs_time)` seconds on a laptop computer. For this particular model, the computation time can be sped up by using an analytic solution to the transition matrix eigenvalues: $\lambda_i = F_i S_{Ji} + S_{Ai}$.
Fitting the analytic version of the model with the same Bayesian inference routine took `r round(bbs_analytic_time)` seconds on the same machine.
Such convenient and computationally efficient analytic solutions will not be available for all DSDM models however - particularly where matrices are large or encode more complex demography, such as trait-dependent or temporally stochastic vital rates, density dependence, or dispersal. Moreover, deriving analytic solutions for quantities relating to matrix population models is likely to be outside the skill-set of most ecologists, and in any case would force users to focus on model implementation details rather than the ecology of their study species. Consequently, the ability to efficiently analyse large numbers of population transition matrices during each iteration of the inference algorithm is critical to ensuring DSDMs are a practically useful tool for ecologists.

>>>>

The spatially-averaged fecundity and adult survival estimates from @ryuDevelopingPopulationModels2016 were used to define priors on regression coefficients for the vital rate submodels. These estimated vital rates correspond to populations at the $J$ survey stations used in the recapture analysis (station names provided in the supplemental data to @ryuDevelopingPopulationModels2016) - locations where the species occurs. I defined a prior on $\beta_F$ and $\beta_S$ such that values were *a priori* more likely if they resulted in estimates of $F_j$ and $S_{A_j}$, for each survey station location $j$, that were more consistent with the estimates in @ryuDevelopingPopulationModels2016.
First, the means and variances of these estimates were used to calculate equivalent parameters of univariate normal distributions on the link-scale of the regression submodels, resulting in the implied priors: $ln(F_j) \sim N(-0.63, 0.18)$, $logit(S_{A_j}) \sim N(1.12, 1.28)$, and $logit(S_{J_j}) \sim N(-0.05, 0.85)$.
These normally-distributed implied priors correspond exactly to multivariate normal priors on the vectors of regression coefficients, since the matrix-multiplication that maps between regression coefficients and link-scale response is an affine transformation, under which the multivariate normal distribution is closed. The parameters of the priors over $\beta_F$ are therefore given by: $\mu_F = \mathbf{X}_*^{-1}(\text{-}0.63\, \mathbf{I})$ and $\mathbf{\Sigma} = \mathbf{X}_*^{-1} (0.18\,\mathbf{I}) \mathbf{X}_*^{-T}$, where $\mathbf{X}_*$ denotes the design matrix over the $J$ survey stations, $\mathbf{X}_*^{-1}$ its Moore-Penrose pseudoinverse, $\mathbf{X}_*^{-T}$ the transpose of that pseudoinverse, and $\mathbf{I}$ an identity matrix of dimension $J$. The prior parameters $\mu_S$ and $\mathbf{\Sigma}_S$ were defined in the same way using the implied prior parameters over $logit(S_{A_j})$: 1.12, and 1.28.
Note that these priors do not provide any information on the direction of the relationships between vital rates and environmental covariates (the intercept elements of $\mu_F$ and $\mu_S$ are informed, but the prior mean for the other terms is $\approx 0$), but by providing information on plausible magnitudes of, and correlations between these parameters the prior favours regression equations that pass close to the vital rate estimates at $F_j$ and $S_j$.
The prior for $\gamma$, the logit-difference between adult and juvenile survival, is given by the difference between the implied normal priors on logit-survival of adults and juveniles, which is also normal.
In the absence of any meaningful prior information for the competition effects or detection rates, I assigned $\alpha$ an improper uniform prior on all positive values.

@@

For example, theese linear models could be applied to very large numbers of covariates and transformations thereof in combination with regularising priors to perform automatic variable selection, similarly to the popular Maxent SDM software[@phillipsModelingSpeciesDistributions2008]. Or to specify mixed-effects models to account for clustered sampling designs, or use random slopes to share environmental reponse information between species in a biotic interaction DSDM. Instead of linear models, non-linear functions could be introduced with generalisd additive models or Gaussian process models, for which there are existing greta extension packages: greta.gam and greta.gp [@goldingGretaGamGeneralised2019; @goldingGretaGpGaussian2019]. These can also be used to model spatial and teemporal random effects and account for drivers of species vital rates that are not captured in the measured environmental variables. Prespecified mathematical relationships or covariate values can also be included, enabling incorporation of  spatial and temporal predictions from physiological models.

@@

Bayesian inference on the DSDM for the Carolina Chickadee by MCMC took `r round(bbs_time)` seconds on a laptop computer. For this particular model, the computation time can be sped up by using an analytic solution to the transition matrix eigenvalues: $\lambda_i = F_i S_{J_i} + S_{A_i}$.
Fitting the analytic version of the model with the same Bayesian inference routine took `r round(bbs_analytic_time)` seconds on the same machine.
Such convenient and computationally efficient analytic solutions will not be available for all DSDM models however - particularly where matrices are large or encode more complex demography, such as trait-dependent or temporally stochastic vital rates, density dependence, or dispersal. Moreover, deriving analytic solutions for quantities relating to matrix population models is likely to be outside the skill-set of most ecologists, and in any case would force users to focus on model implementation details rather than the ecology of their study species. Consequently, the ability to efficiently analyse large numbers of population transition matrices during each iteration of the inference algorithm is critical to ensuring DSDMs are a practically useful tool for ecologists.
